---
title: "DSC1107 Lecture 3"
author: "Lindsay Faith Bazar"
date: "February 02, 2025"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
# Load diamonds dataset
library(ggplot2)
data(diamonds)
head(diamonds)
```

## 2.1: Filtering Diamonds with Ideal Cut and at Least 3 Carats

```{r}
ideal_large_diamonds <- diamonds %>% filter(cut == "Ideal" & carat >= 3)
nrow(ideal_large_diamonds)
```

## 2.2: Selecting All Columns Except x, y, z

```{r}
diamonds_selected <- diamonds %>% select(-x, -y, -z)
head(diamonds_selected)
```

## 2.3: Arranging Diamonds by Decreasing Length

```{r}
diamonds %>% arrange(desc(x)) %>% head(1)
```

## 3.1: Adding a Column for Good Color (D, E, F, G)

```{r}
diamonds <- diamonds %>% mutate(good_color = color %in% c("D", "E", "F", "G"))
head(diamonds)
```

## 3.2: Checking if Any Diamond of at Least 1 Carat Costs Less Than $1000

```{r}
diamonds %>% filter(carat >= 1, price < 1000) %>% summarise(any_below_1000 = n() > 0)
```

## 4.1: Mean Price for Diamonds of Volume at Least 1 Carat

```{r}
diamonds %>% mutate(volume = x * y * z) %>% filter(volume >= 1) %>% summarise(mean_price = mean(price))
```

## 4.2: Counting Diamonds by Cut Using group_by and summarise

```{r}
diamonds %>% group_by(cut) %>% summarise(count = n())
```

## Visualizations

### Histogram of Price Per Carat for Premium Cut Diamonds

```{r}
diamonds %>% 
  filter(cut == "Premium") %>% 
  mutate(price_per_carat = price / carat) %>% 
  ggplot(aes(x = price_per_carat)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  theme_minimal()
```

### Maximum Price for Each Cut

```{r}
diamonds %>% 
  group_by(cut) %>% 
  summarise(max_price = max(price)) %>%
  ggplot(aes(x = cut, y = max_price, fill = cut)) +
  geom_col() +
  theme_minimal()
```

## Exercises

### Minimum Diamond Price (Two Ways)

```{r}
diamonds %>% summarise(min_price = min(price))
diamonds %>% arrange(price) %>% slice(1)
```

### Diamonds with Length At Least 1.5 Times Width

```{r}
diamonds %>% filter(x >= 1.5 * y) %>% summarise(count = n())
```

### Median Carat for Diamonds of Each Cut (Among Colors D, E, F, G)

```{r}
diamonds %>% 
  filter(color %in% c("D", "E", "F", "G")) %>%
  group_by(cut) %>%
  summarise(median_carat = median(carat))
```