---
title: "DSC1107 Lecture 2"
author: "Lindsay Faith Bazar"
date: "February 02, 2025"
output: html_document
---


```{r setup, include=FALSE}
library(tidyverse)
library(ggcorrplot)
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
diamonds %>% glimpse()
```

# Exploratory Data Analysis
### Understanding Variables
1. **Variables**: The dataset has 10 variables (`carat`, `cut`, `color`, `clarity`, `depth`, `table`, `price`, `x`, `y`, `z`).
2. **Observations**: Each observation represents a diamond, with 53,940 diamonds in total.
3. **Values**: Examples include `0.23` for `carat`, `Ideal` for `cut`, `E` for `color`, and `326` for `price`.
4. **Continuous Variables**: `carat`, `depth`, `table`, `price`, `x`, `y`, and `z`.
5. **Categorical Variables**: `cut`, `color`, and `clarity`.

```{r}
summary(diamonds)
```

# Variation
## Discrete Variables
Bar chart of `cut` variation:
```{r}
ggplot(diamonds, aes(x = cut)) + 
  geom_bar(fill = "steelblue") + 
  theme_minimal()
```

3.1. **Most common diamond cut**: `Ideal` is the most common, followed by `Premium`.

## Continuous Variables
Histogram of `carat`:
```{r}
ggplot(diamonds, aes(x = carat)) + 
  geom_histogram(binwidth = 0.1, fill = "steelblue", color = "black") + 
  theme_minimal()
```

3.2. **Variation in `carat`**: Most diamonds have a `carat` value below `3`, indicating that smaller diamonds are more frequent.

# Covariation
## Discrete vs. Discrete
Count plot of `cut` vs `clarity`:
```{r}
ggplot(diamonds, aes(x = cut, y = clarity)) + 
  geom_count() + 
  theme_minimal()
```

4.1. **Relationship between `cut` and `clarity`**: `Ideal` cut diamonds have more occurrences across different clarity levels.

## Continuous vs. Continuous
Scatter plot of `x` vs `y`:
```{r}
ggplot(diamonds, aes(x = x, y = y)) + 
  geom_point(alpha = 0.3) + 
  geom_abline(slope = 1, intercept = 0, color = "red") + 
  theme_minimal()
```

4.2. **Insight**: Most points lie close to `y = x`, indicating that diamonds tend to be symmetric in their length and width.

## Continuous vs. Discrete
Boxplot of `price` by `cut`:
```{r}
ggplot(diamonds, aes(x = cut, y = price)) + 
  geom_boxplot(fill = "steelblue") + 
  theme_minimal()
```

4.3. **Trend**: Higher `cut` quality does not necessarily mean a higher median price.


# Additional Visualization Tools
## Axis Transformations
```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(alpha = 0.3) + 
  scale_x_log10() + 
  scale_y_log10() + 
  theme_minimal()
```

## Multiple Aesthetics
```{r}
ggplot(diamonds, aes(x = carat, y = price, color = cut)) + 
  geom_point(alpha = 0.3) + 
  theme_minimal()
```

## Faceting
```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(alpha = 0.3) + 
  facet_wrap(~cut) + 
  theme_minimal()
```


# Exercises
## 7.1 Relating the Carat and Cut of a Diamond
```{r}
ggplot(diamonds, aes(x = cut, y = carat)) + 
  geom_boxplot(fill = "steelblue") + 
  theme_minimal()
```

**Conclusion**: Higher quality cuts tend to have lower median carat values, likely because larger diamonds are often cut for weight retention rather than optimal proportions.

## 7.2 Relating the Size and Carat of a Diamond
```{r}
ggplot(diamonds, aes(x = carat, y = x * y * z)) + 
  geom_point(alpha = 0.3, color = "steelblue") + 
  theme_minimal() +
  labs(y = "Size (x * y * z)")
```

**Conclusion**: There is a strong positive relationship between carat and size, which means that knowing the carat value gives a fairly accurate idea of the diamond's actual dimensions.
